Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-08-12 review/Response to reviews.Rmd

> L277--278---How does this show the effect of size on growth is greatest for mid-size plants?
> More generally, I got a little confused with the flipping back and forth between growth & plant size in the results and discussion.
> More precise language would help.

We have edited the text to omit "growth", and hopefully clarify: "...the impact of plant size on size in the following year was greatest for mid-sized plants in continuous forest." <!--# are there other edits we can make to clarify?  A sentence introducing the concept that size t+1 ~ size t *is* growth? -->

> L331--332---How can you see in this figure that the response is 'more muted as indicated by the shape of the crossbasis function'?

Our point was actually that SPEI had less of an impact on flowering probability in fragments.
We edited this sentence to read: "SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure 6b)."

@@

> Lines 251ff: I don't see a reference of Fig 2 in the result section.

##  {#R2-5}

> Line 277-278: I don't see in Figure 3a that the impact of size on growth was greatest for mid-sized plants in continuous forest.
> It looks to me that the 1-ha and CF lines are extremely close together and not statistically different.
> Perhaps you can say that there is a tendency of small plants having a larger growth in fragments while large plants grow slightly more in CF.

<!--# I think this will become more clear if I switch to plotting these as back-transformed. See comments in R1-figs -->

@@

> Line 205-207 -- "number of knots not important ..." yes, but package mgcv includes tools for checking those assumptions, would be helpful to include residual analysis and the results of gam.check() in an appendix.

##  {#R3-6}

> Line 211 - random effect -- how did you deal with this in the predictions?
> set to zero in design matrix, Choose a particular value, sample from the posterior?
> The code would tell me but not available.

What we show in the figures are not predictions (i.e. results of `predict()` in R), but evaluated smooths generated with the `gratia` package (equivalent to `mgcv::plot.gam()`).
For figures 4, 5, and 6 we have added the model intercepts to those evaluated smooths and back-transformed (when applicable) the estimates.
These plots, therefore, do not include the additive effects of other smooths, parametric covariates, or random effects.
We think that this is an intuitive way to display the results.
We added a paragraph to the methods explaining our strategy for visualizing the GAM results.

We apologize for not including a link to the code for the project.
All analysis code can be found here: [\<https://github.com/BrunaLab/HeliconiaDemography\>](https://github.com/BrunaLab/HeliconiaDemography){.uri}.
It is so rare for a reviewer to make suggestions based on analysis code, and we are very grateful for these comments!

### 2021-08-12 review/Response to reviews.Rmd

##  {#R3-2}

> Line 166 - given the shallow roots, why not use 1 month SPEI and let the model do the smoothing over longer time frames?
> You describe why not using longer smooths, but not shorter.

##  {#R3-3}

> Line 191 - 194 (equation and description) for clarity you should describe what the index variables i and j are indexing, and I think clearer to have y_ij for ith plant in jth plot.
> Indexes on the random effect description suggest there is a separate variance estimated for each plot?
> d_it suggests that each plant has a separate SPEI estimate?
> I think it is probably aggregated at a higher level than individual?

We thank the reviewer for the very helpful comments on our model equation.
We have changed $y_i$ to $y_{ij}$ and added an explanation of the indexing as suggested.
We have also corrected the error in the random effect equation so that only one variance is estimated.
$d_{it}$ is intended to indicate that each plant observation is associated with a measure of SPEI for the past 36 months.
The SPEI history varies among plants spatially (there are three grid cells in the weather data within BDFFP, so different plots experience different climate) and temporally (observations from all years are pooled).

> I would reorganize the following paragraph so that all the descriptive text for the components of (1) occurs before the digression on the details of the crossbasis function.

We moved the explanation of covariates and random effects earlier as suggested.

### 2021-08-12 doc/paper.Rmd
improve (technical) explanation of models.  Closes #87

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models [DLNMs, @gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recover from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

@@

$$
  \begin{split}
  & g[ \textrm{E}(y_{i})] = \alpha_j + s_1(z_i) + s_2(d_{i,t}, ..., d_{i,t-36}) + \beta x_i \\
  & \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2_{\alpha_j}), \;\textrm{for plot j} 
  \end{split}
$${\#eq:eqn1}

where $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details).
The crossbasis function, $s_2(\cdot)$ can also be written:

@@

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see chapter 5 of @woodGeneralizedAdditiveModels2017 and @gasparriniPenalizedFrameworkDistributed2017 for more detail).
Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of lagged SPEI values for each observation with 36 months as a maximum lag) and 3 knots for the drought response dimension to restrict the shape of the fitted response to drought to bimodal when most complex.
Because of penalization, the number of knots is generally not important as long it is large enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.

To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution because residuals were leptokurtic with a Gaussian error structure.
Because number of inflorescences was highly zero-inflated, we converted this to a binary response to model reproduction (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
Additionally, in models for flowering probability and size, we included plant ID as a random effect to account for variation among individuals.
Preliminary analyses showed that this random effect was not significant in the growth models (edf \~ 0, p \> 0.2) and as such it was dropped to improve computational efficiency.
A random effect of plant ID was not included in survival models since each plant only dies once.

>>>>

To assess the effects of drought history on plant vital rates we used Distributed Lag Non-linear Models [DLNMs, @gasparriniPenalizedFrameworkDistributed2017].
DLNMs capture how potentially delayed effects of predictor variables (e.g. SPEI) affect an outcome (e.g. growth) well beyond the event period.
<!--# DLNMs solve the problems created by including all possible climate windows in a single model---namely collinearity of predictors and the arbitrary choice of window length -->They do so by fitting a bi-dimensional predictor-lag-response association spline, referred to as a crossbasis function.
This models a non-linear relationship between predictor and response (e.g. between SPEI and vital rates) and allows the shape of that relationship to vary smoothly over lag time.
<!--# The shape of the function can be penalized so that it is only as complex as the data supports, thus reducing effective degrees of freedom --> Using the `dlnm` package [@gasparriniDistributedLagLinear2011; @rcoreteamLanguageEnvironmentStatistical2020], we created crossbasis functions with possible lags from 0--36 months.
We chose 36 months as a maximum lag because prior transplant experiments with *H. acuminata* showed they typically recover from transplant shock in less than 36 months [@brunaExperimentalAssessmentHeliconia2002] so this is a reasonable upper bound for lagged effects of drought.

@@

$$
  \begin{split}
  & g[ \textrm{E}(y_{ij})] = \alpha_j + s_1(z_i) + s_2(d_{i,t}, ..., d_{i,t-36}) + \beta x_i \\
  & \alpha_j \sim N(\mu_{\alpha_j}, \sigma^2), \;\textrm{for plot j} 
  \end{split}
$${\#eq:eqn1}

where $y_{ij}$ is the $i$th plant observation in the $j$th plot $s_1(z_i)$ is a smooth function of plant size (natural log of height $\times$ shoot number), fit using a penalized cubic regression spline, $s_2(\cdot)$ is the crossbasis function in which $d_{i,t}$ is the SPEI value during the census month of an observation (February) and $d_{i,t-l}$ is the SPEI $l$ months prior (see Gasparrini et al. 2017 for details).
To determine if plot characteristics influenced average vital rates we included a random effect of plot ID on the intercept; this was represented by $a_j$ in @eq:eqn1.
We modeled a potential cost of reproduction by including flowering in the previous year as covariate, $x_i$, in @eq:eqn1.
Additionally, in models for flowering probability and size, we included plant ID as a random effect to account for variation among individuals.
Preliminary analyses showed that this random effect was not significant in the growth models (edf \~ 0, p \> 0.2) and as such it was dropped to improve computational efficiency.
A random effect of plant ID was not included in survival models since each plant only dies once.
The crossbasis function, $s_2(\cdot)$ can also be written:

@@

where the crossbasis function, $f \cdot w (d,l)$, is composed of two marginal basis functions: the standard predictor-response function $f(d)$, and the additional lag-response function $w(l)$.
These marginal functions are combined as a tensor product smooth such that the shape of one marginal function varies smoothly along the other dimension (see chapter 5 of @woodGeneralizedAdditiveModels2017 and @gasparriniPenalizedFrameworkDistributed2017 for more detail). Penalized cubic regression splines were used for both marginal bases of the crossbasis function, with 35 knots for the lag dimension (i.e. number of lagged SPEI values for each observation with 36 months as a maximum lag) and 3 knots for the drought response dimension to restrict the shape of the fitted response to drought to bimodal when most complex. Because of penalization, the number of knots is generally not important as long it is large enough to allow the smooth to represent the 'true' relationship [@woodGeneralizedAdditiveModels2017].
<!--# reference supplemental model diagnostics here --> Estimated degrees of freedom (edf) represent the 'true' complexity of the smooth after penalization with edf = 1 being equivalent to a straight line and larger numbers representing more complex curves.

We determined the effects of SPEI on plant growth using plant size in year t+1 as a response variable.
This was modeled with a scaled t family error distribution because residuals were leptokurtic with a Gaussian error structure.
Because number of inflorescences was highly zero-inflated, we converted this to a binary response to model reproduction (i.e., 1 for ≥1 inflorescence, 0 for no inflorescences).
We modeled both reproduction and survival (i.e., from year t to year t+1) using a binomial family error distribution with a logit link function.
